---
title: "Word Cloud"
output: html_notebook
---



```{r echo=FALSE}
# http://www.sthda.com/english/wiki/text-mining-and-word-cloud-fundamentals-in-r-5-simple-steps-you-should-know
# Install
# install.packages("tm")  # for text mining
# install.packages("SnowballC") # for text stemming
# install.packages("wordcloud") # word-cloud generator 
# install.packages("RColorBrewer") # color palettes
# Load
library("tm")
library("SnowballC")
library("wordcloud")
library("RColorBrewer")
```

```{r}
setwd("~/Google Drive/Classroom/Economía del Bienestar/2020")
archivos<- list.files("~/Google Drive/Classroom/Economía del Bienestar/2020")
text <- readLines(paste("/Users/sebacea/Google Drive/Classroom/Economía del Bienestar/2020/",archivos[1],sep=""))
for (archivo in 2:length(archivos)) {
  text <- paste(text,readLines(paste("/Users/sebacea/Google Drive/Classroom/Economía del Bienestar/2020/",archivos[archivo],sep="")))
}
text <- readLines(file.choose())
docs <- Corpus(VectorSource(text))
inspect(docs)
```

Transformation is performed using tm_map() function to replace, for example, special characters from the text.

Replacing “/”, “@” and “|” with space:

```{r}
toSpace <- content_transformer(function (x , pattern ) gsub(pattern, " ", x))
docs <- tm_map(docs, toSpace, "/")
docs <- tm_map(docs, toSpace, "@")
docs <- tm_map(docs, toSpace, "\\|")
```

The R code below can be used to clean your text :
```{r}
# Convert the text to lower case
docs <- tm_map(docs, content_transformer(tolower))
# Remove numbers
docs <- tm_map(docs, removeNumbers)
# Remove english common stopwords
docs <- tm_map(docs, removeWords, stopwords("english"))
docs <- tm_map(docs, removeWords, stopwords("spanish"))
# Remove your own stop word
# specify your stopwords as a character vector
docs <- tm_map(docs, removeWords, c("fundación", "objetivo", "proyecto", "programa", "través")) 
# Remove punctuations
docs <- tm_map(docs, removePunctuation)
# Eliminate extra white spaces
docs <- tm_map(docs, stripWhitespace)
# Text stemming
# docs <- tm_map(docs, stemDocument)
```

Document matrix is a table containing the frequency of the words. Column names are words and row names are documents. The function TermDocumentMatrix() from text mining package can be used as follow :

```{r}
dtm <- TermDocumentMatrix(docs)
m <- as.matrix(dtm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
head(d, 10)
```

The importance of words can be illustrated as a word cloud as follow :

```{r, echo=FALSE}
set.seed(1234)
wordcloud(words = d$word, freq = d$freq, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))
```

# Base datos Colunga

```{r}
library(readxl)
Base_de_Datos_Colunga <- read_excel("Colunga/Base de Datos Colunga.xlsx")
View(Base_de_Datos_Colunga)
```